Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2018-02-15 RNARedPrint-Bioinfo2018.tex
cleaned up biblio + acknowledgements

A further \Def{rejection step} is applied to the generated structures to retain only sequences whose energy for each structure $R_\ell$ belongs to $[\TargetE_\ell\cdot(1-\varepsilon),\TargetE_\ell\cdot(1+\varepsilon)]$, for $\varepsilon\ge 0$ some predefined \Def{tolerance}. The rationale for a rejection approach follows from: 
i) \emph{Enacting an exact control over the energies would  be technically hard and costly.} Indeed, controlling the energies through dynamic programming would require explicit convolution products, generalizing~\citet{Cupal1996}, inducing additional $\Theta(n^{2k})$ time and $\Theta(n^k)$ space overheads;
ii) \emph{Induced distributions are typically concentrated.} Intuitively, unless sequences are fully constrained individual energy terms are independent enough so that their sum is concentrated around its mean -- the targeted energy (cf Fig.~\ref{fig:energydist}). 
%Indeed, as shown in Fig.~\ref{fig:shifting_mean}, empirical joint distributions in the energies have good fits towards Normal multidimensional distributions with linear (co-)variances. 
For base pair-based energy models and special compatibility graph
(paths, cycles\ldots) this property rigorously follows from analytic
combinatorics, see Bender-Richmond-Williamson~\citep{Bender1983} and
Drmota~\citep{Drmota1997}. In such cases, the expected number of
rejections before reaching the targeted energies remains constant when
$\varepsilon\ge 1/\sqrt{n}$, and $\Theta(n^{k/2})$ when
$\varepsilon=0$. The \GCb-content of designs can also be controlled,
jointly and in a similar way, as done in
\Software{IncaRNAtion}~\citep{Reinharz2013}.

@@

To illustrate our above strategy, we sampled $n=1\,000$ sequences
targeting ${\sf GC}_{target}=0.5$ and different Turner energies for
the three structure targets of an example
instance. Fig.~\ref{fig:energydist}A illustrates how the Turner
energy distributions of the shown target structures can be accurately
shifted to prescribed target energies; for comparison, we plot the
respective energy distribution of uniformly and Boltzmann sampled
sequences. Fig.~\ref{fig:energydist}B summarizes the targeting accuracy for the single structure energies over a larger set of instances
from the \texttt{Modena} benchmark. We emphasize that only the simple stacking energies
are directly targeted (at $\pm$10\% tolerance, expect for a few hard
instances). Due to our linear adjustment, we still achieve well
controlled energies in the ultimately relevant Turner energy model.
\begin{figure*}[t]
  \begin{center}
    {\sf \bfseries A}\includegraphicstop[width=0.57\textwidth]{Figs/energy_distribution}
    {\sf \bfseries B}\includegraphicstop[width=0.32\textwidth]{Figs/offset}
  \end{center}
  \caption{%
  	Targeting specific energies using multi-dimensional Boltzmann sampling. (A) Turner energy distributions for three target structures (right) while targeting $(-40,-40,-20)$ and $(-20,-20,-20)$ free energy vectors. For comparison, we show the distribution of uniform and Boltzmann samples; respectively associated with homogenous weights $1$ and $e^\beta$. (B) Accuracy of targeting over all target structures of the \Software{Modena} benchmark instances \texttt{RNAtabupath} (2str), \texttt{3str} and \texttt{4str}. Shown is the relative deviation of targeted and achieved energies in the simple stacking model and the Turner model.
%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
}
  \label{fig:energydist}
\end{figure*}
\subsection{Generating high-quality seeds for further optimization}
We empirically study the capacity of \ourprog{} to improve the generation of seed sequences for subsequent local optimizations, showing an important application in biologically relevant scenario.

@@

\section*{Acknowledgements}
YP is supported 
%jointly supported 
by
% the French 
{\em Agence
  Nationale de la Recherche} and the 
%Austrian
FWF (ANR-14-CE34-0011; project RNALands).  SH is supported by the
German Federal Ministry of Education and Research (BMBF support code
031A538B; de.NBI: German Network for Bioinformatics Infrastructure)
and the
% acknowledged by
%financial support 
of the Future and Emerging Technologies programme
% within the Seventh Framework Programme for Research of the
%European Commission, under 
(FET-Open grant 323987; project RiboNets).
%
We thank Leonid Chindelevitch for pointing out the potential for
complexity reductions due to the bipartiteness of the compatibility
graph; moreover, Arie Koster for discussing the practical computation
of the tree decomposition.

>>>>

A further \Def{rejection step} is applied to the generated structures to retain only sequences whose energy for each structure $R_\ell$ belongs to $[\TargetE_\ell\cdot(1-\varepsilon),\TargetE_\ell\cdot(1+\varepsilon)]$, for $\varepsilon\ge 0$ some predefined \Def{tolerance}. The rationale for a rejection approach follows from: 
i) \emph{Enacting an exact control over the energies would  be technically hard and costly.} Indeed, controlling the energies through dynamic programming would require explicit convolution products, generalizing~\citet{Cupal1996}, inducing additional $\Theta(n^{2k})$ time and $\Theta(n^k)$ space overheads;
ii) \emph{Induced distributions are typically concentrated.} Intuitively, unless sequences are fully constrained individual energy terms are independent enough so that their sum is concentrated around its mean -- the targeted energy (cf Fig.~\ref{fig:energydist}). 
%Indeed, as shown in Fig.~\ref{fig:shifting_mean}, empirical joint distributions in the energies have good fits towards Normal multidimensional distributions with linear (co-)variances. 
For base pair-based energy models and special compatibility graph
(paths, cycles\ldots) this property rigorously follows from analytic
combinatorics, see \citet{Bender1983} and
\citet{Drmota1997}. In such cases, the expected number of
rejections before reaching the targeted energies remains constant when
$\varepsilon\ge 1/\sqrt{n}$, and $\Theta(n^{k/2})$ when
$\varepsilon=0$. The \GCb-content of designs can also be controlled,
jointly and in a similar way, as done in
\Software{IncaRNAtion}~\citep{Reinharz2013}.

@@

To illustrate our above strategy, we sampled $n=1\,000$ sequences
targeting ${\sf GC}\%=0.5$ and different Turner energies for
the three structure targets of an example
instance. Fig.~\ref{fig:energydist}A illustrates how the Turner
energy distributions of the shown target structures can be accurately
shifted to prescribed target energies; for comparison, we plot the
respective energy distribution of uniformly and Boltzmann sampled
sequences. Fig.~\ref{fig:energydist}B summarizes the targeting accuracy for the single structure energies over a larger set of instances
from the \texttt{Modena} benchmark. We emphasize that only the simple stacking energies
are directly targeted (at $\pm$10\% tolerance, expect for a few hard
instances). Due to our linear adjustment, we still achieve well
controlled energies in the ultimately relevant Turner energy model.
\begin{figure*}[t]
  \begin{center}
    {\sf \bfseries A}\includegraphicstop[width=0.57\textwidth]{Figs/energy_distribution}
    {\sf \bfseries B}\includegraphicstop[width=0.32\textwidth]{Figs/offset}
  \end{center}
  \caption{%
  	Targeting specific energies using multi-dimensional Boltzmann sampling. (A) Turner energy distributions for three target structures (right) while targeting $(-40,-40,-20)$ and $(-20,-20,-20)$ free energy vectors. For comparison, we show the distribution of uniform and Boltzmann samples; respectively associated with homogenous weights $1$ and $e^\beta$. (B) Accuracy of targeting over all target structures of the \Software{Modena} benchmark instances \texttt{RNAtabupath} (2str), \texttt{3str} and \texttt{4str}. Shown is the relative deviation of targeted and achieved energies in the simple stacking model and the Turner model.
%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
}
  \label{fig:energydist}
\end{figure*}
\subsection{Generating high-quality seeds for further optimization}
We empirically study the capacity of \ourprog{} to improve the generation of seed sequences for subsequent local optimizations, showing an important application in biologically relevant scenario.

@@

\section*{Acknowledgements}
YP is supported 
%jointly supported 
by the
% French 
{\em Agence
  Nationale de la Recherche} and the 
%Austrian
FWF (ANR-14-CE34-0011; project RNALands).  SH is supported by the
German Federal Ministry of Education and Research (BMBF support code
031A538B; de.NBI: German Network for Bioinformatics Infrastructure)
and the
% acknowledged by
%financial support of the 
Future and Emerging Technologies programme
% within the Seventh Framework Programme for Research of the
%European Commission, under 
(FET-Open grant 323987; project RiboNets).
%
We thank Leonid Chindelevitch for suggesting using the stable sets analogy to optimize our FPT algorithm, and Arie Koster for practical recommendations 
for computing tree decompositions.
